### 2025-10-27 Text/Thesis/Chapters/Chapter1.qmd
final version thesis

Synthetic Aperture Radar (SAR) technology has become essential in  remote sensing, as it offers high-resolution imaging capabilities independently of sunlight and weather conditions, thus facilitating continuous and reliable Earth observation&nbsp;[@Mondini2021; @Zeng2020]. SAR imagery finds extensive applications in environmental monitoring, disaster management, agricultural mapping, climatology, urban planning, and maritime surveillance, among others&nbsp;[@Akbarizadeh2012;@Moreira2013; @Mu2019;@Yu2023]. Despite these advantages, effective exploitation of SAR data requires a robust understanding of its unique statistical characteristics, particularly the presence of speckle noise, a granular interference pattern inherent to coherent imaging systems like SAR&nbsp;[@Argenti2013;@Choi2019].

@@

Parameter estimation for these statistical models, however, presents significant challenges, especially when using small-sized windows necessary to preserve spatial resolution in SAR images. Traditional likelihood-based or moment-based estimators often become biased or unstable under these practical constraints&nbsp;[@Nascimento2010; @VasconcellosFrerySilva]. This limitation motivates alternative statistical approaches not reliant on precise parameter estimation or large sample sizes.

@@

<!-- The primary objective of this thesis is to develop and validate a set of entropy-based statistical tests for the detection of heterogeneity in SAR intensity images. These tests are designed to distinguish between fully developed speckle, modeled by the Gamma distribution, and heterogeneous clutter, modeled by the $\mathcal{G}_I^0$ distribution, in a robust and interpretable manner. -->

@@

<!-- 4. **Comprehensive validation.** Evaluation of the statistical performance of the proposed tests through extensive Monte Carlo simulations, analyzing Type I error (size), power, and robustness under various speckle and texture conditions. -->

<!-- 5. **Real applications.** Application of the proposed methodology to both simulated and real SAR data, producing $p$-value maps through sliding window analysis for visual and quantitative assessment of spatial heterogeneity. -->

## COMPUTATIONAL SUPPORT AND REPRODUCIBILITY

Simulations, tests, and data analyses presented in this thesis were implemented using the free, open-source programming language \texttt{R}\footnote{\url{https://www.r-project.org/}} within \texttt{RStudio} (version 2024.12.1+563, \texttt{R} version 4.4.2). The manuscript was written in \texttt{Quarto}\footnote{\url{https://quarto.org/}}, a dynamic publishing system that integrates code, output, and narrative to ensure full reproducibility&nbsp;[@bauer2023reproducible].

The numerical experiments were performed on a personal computer equipped with an AMD Ryzen 9 5900X 12-Core Processor (3.70 GHz) and 32 GB RAM.

@@

The following peer-reviewed articles and conference paper have been published as part of or related to this research:

1. [@Alpala2025]: [Quantifying Heterogeneity in SAR Imagery with the Rényi Entropy.](https://openaccess.wgtn.ac.nz/articles/journal_contribution/Quantifying_Heterogeneity_in_SAR_Imagery_with_the_R_nyi_Entropy/29376944?file=55548095)

2. [@Frery2024]: [Identifying Heterogeneity in SAR Data with New Test Statistics.](https://www.mdpi.com/2072-4292/16/11/1973)

3. [@Alpala2024]: [Identifying departures from the fully developed speckle hypothesis in intensity SAR data with non-parametric estimation of the entropy.](https://ieeexplore.ieee.org/document/10544448)

>>>>

Synthetic Aperture Radar (SAR) technology has become essential in  remote sensing, as it offers high-resolution imaging capabilities independently of sunlight and weather conditions, thus facilitating continuous and reliable Earth observation&nbsp;[@Mondini2021]. 
SAR images are widely applied in environmental monitoring&nbsp;[@Amitrano2021], 
disaster assessment&nbsp;[@Ge2020], 
agriculture&nbsp;[@Changan2019], 
urban planning&nbsp;[@Esch2010;@NovelTechniquesforBuiltupAreaExtractionfromPolarimetricSARImages2019], and  maritime surveillance&nbsp;[@Cao2024;@OntheDetectionandLongTermPathVisualisationofa68Iceberg], among other fields&nbsp;[@Moreira2013].
Despite these advantages, effective exploitation of SAR data requires a robust understanding of its unique statistical characteristics, particularly the presence of speckle noise, a granular interference pattern inherent to coherent imaging systems like SAR&nbsp;[@Argenti2013].

@@

Parameter estimation for these statistical models, however, presents significant challenges, especially when using small-sized windows necessary to preserve spatial resolution in SAR images.
Traditional parametric approaches, such as maximum likelihood estimation and moment-based methods, often become biased or unstable under these practical constraints&nbsp;[@Nascimento2010; @VasconcellosFrerySilva]. Moreover, likelihood ratio tests derived from these estimators may exhibit unreliable behavior when applied to small-window SAR data due to flat likelihood surfaces.

@@

@@

The following peer-reviewed articles and conference papers have been published as part of or related to this research:

- [-@Alpala2025]: [ALPALA, J.; NASCIMENTO, A. D. C.; FRERY, A. C. Quantifying Heterogeneity in SAR Imagery with the Rényi Entropy. \textit{IEEE Geoscience and Remote Sensing Letters}, p. 1--1, 2025.](https://openaccess.wgtn.ac.nz/articles/journal_contribution/Quantifying_Heterogeneity_in_SAR_Imagery_with_the_R_nyi_Entropy/29376944?file=55548095)

- [-@Frery2024]: [FRERY, A. C.; ALPALA, J.; NASCIMENTO, A. D. C. Identifying Heterogeneity in SAR Data with New Test Statistics. \textit{Remote Sensing}, v. 16, n. 11, p. 1973, 2024.](https://www.mdpi.com/2072-4292/16/11/1973)

- [-@Alpala2024]: [ALPALA, J.; NASCIMENTO, A. D. C.; FRERY, A. C. Identifying departures from the fully developed speckle hypothesis in intensity SAR data with non-parametric estimation of the entropy. In: \textit{2024 International Conference on Machine Intelligence for GeoAnalytics and Remote Sensing (MIGARS)}, 2024.](https://ieeexplore.ieee.org/document/10544448)

- [-@Alpala2023]: [ALPALA, R. J.; BORBA, A. A. De; FRERY, A. C. Quality Assessment measures for explainable fusion Of statistical evidences of edges in Polsar images: A First Approach. In: \textit{IGARSS 2023 - IEEE International Geoscience and Remote Sensing Symposium}, 2023.](https://ieeexplore.ieee.org/abstract/document/10283263)

## COMPUTATIONAL SUPPORT AND REPRODUCIBILITY {#sec-repr}

Simulations, tests, and data analyses presented in this thesis were implemented using the free, open-source programming language \texttt{R}\footnote{\url{https://www.r-project.org/}} within \texttt{RStudio} (version 2025.09.1+401, \texttt{R} version 4.5.1). The manuscript was written in \texttt{Quarto}\footnote{\url{https://quarto.org/}}, a dynamic publishing system that integrates code, output, and narrative to ensure full reproducibility&nbsp;[@bauer2023reproducible].

The numerical experiments were performed on a workstation equipped with an Intel\textsuperscript{\textregistered} Core\texttrademark Ultra 9 285 CPU (2.5 GHz, 24 cores), 128 GB RAM, and an NVIDIA\textsuperscript{\textregistered} GeForce RTX 5090 GPU (31 GB VRAM), running Windows 11 Pro (version 24H2).

@@
